Quantum Bogosort

### 2022-09-06 paper/main.tex
don't steal my stuff until ISEF is over 😭
major post-ISEF cleanup
transitioned to Git LFS hosting of images/PDFs, added ISEF materials,
and updated LaTeX source. I think testing.ipynb has some changes on a
different machine from when I added density matrix calculations...

\title{Quantum Bogosort}
\author{George Huebner}
\date{March 18, 2022}

\begin{document}
\pagenumbering{gobble}

\begin{titlepage}
\maketitle
\centering
\large Chicago STEM Fair, Computer Science Research, Design Category \\
\vspace{3em}
\huge Sponsor \\
\vspace{1em}
\large Michael Caines \hspace{15em}
\end{titlepage}

@@

\pagestyle{fancy}
\fancyhead[R]{\textit{Quantum Bogosort}, Huebner}
\fancyhead[L]{}

@@

\begin{abstract}
\noindent Gag algorithms are a staple of computer science and provide many a good chuckle. Take, for instance, MiracleSort\textsuperscript{\cite{thompson_2013}}, an algorithm that relies on alpha particle emission to cause erroneous bit flips to sort a list:
\begin{minted}{python}
while isSorted(unsorted_list) is False:
    time.sleep(1000)
\end{minted}
Despite their apparent uselessness, joke algorithms provide valuable insight into algorithm design and complexity\textsuperscript{\cite{gruber_holzer_ruepp_2007}}. In this paper we propose an implementation to Quantum Bogosort, one such joke algorithm.
\end{abstract}

@@

Classic Bogosort is a joke sorting algorithm where a list is randomly permuted and checked to see if it is sorted. Essentially, in pseudocode:

@@

\noindent It has an average time complexity of $ O((n+1)!) $ and the nondeterministic (traditional) method has an unbounded worst case time complexity for any non-trivial list.

Quantum Bogosort\textsuperscript{\cite{ryder_2018}}, then, is a variation on classical bogosort. Based on the \href{https://www.pbs.org/wgbh/nova/manyworlds/pdf/dissertation.pdf}{many worlds} interpretation of quantum mechanics, QBS proposes that a quantum computer can superimpose all possible permutations of a list and destroy any universe in which a sorted list does not exist, thus sorting the list in $ O(1) $ time. Because parallel universes are non-communicating, there's no way to know the list is sorted other than by the principle of quantum necessity. The algorithm that satisfies subscribers of the Copenhagen interpretation (where wave functions collapse) is essentially the same thing, but the quantum circuit must be run until the list is sorted.

@@

\section{Background}
Most people are likely unfamiliar with quantum computing and its associated quirks, so a minor detour is in order to explain the technologies that will be used to test the QBS algorithm. Explaining the quantum mechanics that underlie this project is well beyond the scope of a math symposium, though, so let the physics enthusiast proceed carefully.

\subsection{Quantum Computers}
Quantum computers should be thought of not as processors that parse instructions, but rather circuits composed of gates that operate on a group of qubits (\textbf{qu}antum \textbf{bit}, the quantum unit of information). While revolutionary, quantum computers only achieve speedups in certain applications, and have not yet cleared a benchmark of true quantum supremacy due to current hardware implementations. Current noisy intermediate scale quantum (NISQ) computers are highly error prone and have comparatively few qubits and are better thought of as a separate processing unit to be orchestrated by a classical computer.

\subsection{Qubits}
Qubits have quantum properties of superposition, entanglement, and interference that distinguish them from classical bits. Whereas a classical bit is either 1 or 0, a qubit is in a superposition of 1 and 0, which are defined as the following kets:
$$ \ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \text{ (Classical 0)} $$
$$ \ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix} \text{ (Classical 1)} $$
Qubit superpositon can be thought of as having a certain amount (amplitude) of 'one-ness' and 'zero-ness' that can be represented as so:
$$ \ket{x} = \alpha \ket{0} + \beta \ket{1} \ | \ \text{ Probability}_{\ket{0}} = | \alpha |^2 \text{, } \text{ Probability}_{\ket{1}} = | \beta |^2 $$
%Note that $|x|$ here represents magnitude as $\alpha$ and $\beta$ are imaginary.
To actually make use of qubits, they must be measured, which collapses them to either $\ket{0}$ or $\ket{1}$. This is a probabilistic, quantum event, hence the probability\textsubscript{$\ket{x}$}. It is helpful to consider the \hyperref[fig:bsphere]{Bloch sphere} when analyzing qubit state and rotation around the Pauli axes; essentially, the 'North Pole' is defined as $\ket{0}$ and the 'South Pole' is defined as $\ket{1}$. The 'equator,' then, would be an even superposition of $\ket{0}$ and $\ket{1}$.

@@

\noindent Here are a few common qubit gates:
$$ \text{X}(\ket{0}) = \ket{1} $$
$$ \text{X}(\ket{1}) = \ket{0} $$
$$ \text{Z}(\ket{0}) = \ket{0} $$
$$ \text{Z}(\ket{1}) = -\ket{1} $$
$$ \text{H} = \text{R}_{\text{y}} (\frac{\pi}{2}) \text{Z} \ \ \ \text{(Hadamard gate)}$$
$$ \text{H}(\ket{0}) = \frac{1}{\sqrt{2}} \ket{0} + \frac{1}{\sqrt{2}} \ket{1} = \ket{+} $$
$$ \text{H}(\ket{1}) = \frac{1}{\sqrt{2}} \ket{0} - \frac{1}{\sqrt{2}} \ket{1} = \ket{-} $$
Notice that R\textsubscript{y} is an arbitrary rotation around the Pauli Y-axis; Similarly, R\textsubscript{x} and R\textsubscript{z} allow arbitrary rotations along the X- and Z-axes, respectively.\\
Multiple qubits can be used together for entanglement and amplitude interference, and a group of qubits is called a \textit{register}. Although a register can be simplified into a single tensor by taking the tensor product of the qubits in the register, keeping registers in kets makes the math much simpler. \\
Bra-ket (Dirac) notation is used as Kronecker products quickly become unwieldy when dealing with large qubit registers.

@@

\subsection{Entanglement}
Qubit gates can also be controlled if another qubit is $\ket{1}$ (similar to an 'if' clause in a classical programming language). This allows for multiple qubits to be entangled with each other, essentially tying the 'fate' of the qubits together so that they can no longer be thought of as individual qubits but rather a single system. Measuring one qubit will also collapse the other entangled qubits, even if they were never observed. A famous example of this is Schr\"odinger's cat. In this thought experiment, the quantum event of a radioactive substance decaying is entangled with the fate of the cat. If the cat dies, the radioactive substance must have decayed, and if the radioactive substance did not decay, the cat must be alive (or, at the very least, not killed by poison).

@@

\item The other approach explored is essentially a quantum random number generator. Instead of storing the list, we create $ N = l! $ evenly balanced register states ($l!$ being the number of permutations of $array$). A set of states $ [0, N) \cap \mathbb{N} $ will be prepared, each with an even amplitude. Concisely:
    $$ \sum_{x=0}^{N-1} \frac{1}{\sqrt{N}} \ket{x_{BE}} $$
Note that the quantum register will always be represented in \href{https://en.wikipedia.org/wiki/Endianness}{big-endian}.

Because we're storing our number in binary, we want the power of 2 greater than or equal to $ l! $. This means we have $ O(\lceil \log_2 l! \rceil) $ space complexity, which seems scary at first. However, with present-day NISQ technology, this is significantly more favorable than $ O(l \cdot s) $ for any useful data type. Also keep in mind that because element size isn't a concern, any size element or even elements of different types could be stored in the list (as long as every type used can be compared to each other).

@@

\section{Quantum Bogosort}
\subsection{QRNG Algorithm}
The main algorithm is recursive, essentially moving through the register from most to least significant qubit, performing a R\textsubscript{y} gate to split up a qubit into P 'parts,' and entangling the remaining qubits with controlled Hadamard (CH) gates. \\
Let's use $i = 77$ as an example. Recall that for storing an integer $i$, the most significant bit (MSB) used is $\lfloor \log_2 i \rfloor$: For example, the MSB of 77 would be $\lfloor \log_2 77 \rfloor = 64$. Because QRNG generates the integers $[0, 77)$ (excluding 77 itself in order to generate 77 integers), we know that $[0, 64)$ is also being generated, because $[0, 64) \subseteq [0, 77)$. \\
This means that every possible state of the qubit register with a leading zero (in this case, in the 64s place) should be generated. That's already really easy to do! Just use the leading zero as a control and apply Hadamard gates to the rest of the register. Here's a simple example:
\begin{align*}
& \ket{\psi} = \ket{0000000} \\
& H(\psi_0), \ \ \ \ \ \ \ket{\psi} = \frac{1}{\sqrt{2}} \ket{0000000} + \frac{1}{\sqrt{2}} \ket{1000000} \\
& CH(\psi_0, \psi_{[1, 6]}), \ X(\psi_0), \ \ \ \ \ \ \ \ket{\psi} = \frac{1}{\sqrt{128}}(\ket{0000000} + \ket{0000001} + ... + \ket{0111111}) + \frac{1}{\sqrt{2}} \ket{1000000}
\end{align*}
Note that CH is a singly controlled Hadamard applied to multiple qubits, not a multi-controlled Hadamard applied to one qubit. Also note the X gate applied to the MSB because we really want an anti-controlled, not a normal controlled gate. \\
The problem is that we've unevenly split the integer 77 into 2: $|[0, 64)| \neq |[64, 77)|$. Numbers in the range $[64, 77)$ have a $(\frac{1}{\sqrt{13}} \cdot \frac{1}{\sqrt{2}})^2$ probability of being measured whereas numbers in the $[0, 64)$ range have a $(\frac{1}{\sqrt{64}} \cdot \frac{1}{\sqrt{2}})^2$ probability of being measured. \\
To fix this, we have to change the $\frac{1}{\sqrt{2}}$ from the Hadamard gate to something that gives us a more complex superposition; to do this we use the arbitrary rotation gates. $X$ and $Y$ gates give us the desired rotations we want, but we chose $Y$ so we don't have to worry about phase (the sign of a qubit). Phase doesn't matter in this case because only the magnitudes effect the probability of measurement, but it might be important if using QRNG in another algorithm. \\
Superpositions can be represented using spherical coordinates. For example, on the \hyperref[fig:bsphere2]{Bloch sphere}, the spherical coordinates $(1, 0.955, 0)$ would represent the state $\sqrt{\frac{1}{3}} \ket{0} + \sqrt{\frac{2}{3}} \ket{1}$ because $\cos (0.955) \approx \sqrt{\frac{1}{3}}$. We can do simple algebra to get a $\theta$ for an arbitrary superposition. For QRNG, this looks like $\theta = 2\arccos(\sqrt{\frac{77 - MSB}{77}})$. \\
\begin{figure}[H]
    \centering
    \capstart
    \includegraphics[width=0.19\textwidth]{images/bloch_sphere2.png}
    \caption{Bloch sphere\textsuperscript{\cite{wikipedia_2021}}}
    \label{fig:bsphere2}
\end{figure}
\noindent Recall that we have to apply a $X$ gate because gates are controlled by $\ket{1}$, not $\ket{0}$. We can save an extra gate by doubling our rotation (hence the 2), which gives us $\sqrt{\frac{77 - MSB}{77}} \ket{0}$ instead, which flips right back to $\ket{1}$ with another $X$ gate. \\
\indent We then recursively repeat this process, adding the MSB as a control qubit to the R\textsubscript{y} and CH gates and using the 2nd most significant bit as the new MSB. Eventually we split all the states evenly and we're done. \\

\noindent The differences between what is conceptually happening and what is actually implemented are shown below. $\bullet$ is a control and $\circ$ is an anti-control (controls on $\ket{0}$ instead of $\ket{1}$). Note that H\textsubscript{ALL} is \textbf{not} a multi-qubit gate --- it's illustrated that way for clarity.
\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{images/circuits/conceptual_circuit.pdf}
    \caption{Conceptual circuit}
    \vspace{10mm}
    \includegraphics[width=\textwidth]{images/circuits/optimized_circuit.pdf}
    \caption{Circuit with proper controls}
    \label{fig:circs}
\end{figure}

\vspace{20mm}
\noindent Minor implementation details:
\begin{itemize}
    \item Multi-controlled Hadamard (MCH) gates are implemented manually with an ancilla qubit and multi-controlled X (MCX) gates. This could almost certainly be optimized more on a case-by-case basis (i.e. running \mintinline{python}{mode='noancilla'} for machines with fewer qubits).
    \item Especially when there are several qubits in the control register (and therefore the MCH gates are more costly), the MCX on the ancilla doesn't have to be reversed; instead, applying a  \mintinline{python}{'reset'} gate to reset the ancilla to 0 might be more optimal.
    \item \mintinline{python}{circuit} and \mintinline{python}{ancilla} don't change between recursions and needlessly add on to the stack frame of the function. Passing them as parameters is intended for clarity, but if you were to squeeze every drop of performance out they should be made global.
\end{itemize}
\subsubsection{Code}
\begin{minted}{python}
from qiskit import *
from math import acos, sqrt

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{images/example_circuit.png}
    \caption{Circuit diagram for 15 'parts'}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.75\textwidth]{images/example_measurement.png}
    \caption{Measurement results from Aer Simulator}
    \label{fig:measures}
\end{figure}

\subsection{Practicality Assessment}
Obviously bogosort is not meant to be a practical algorithm, but nevertheless we will conduct a review of the efficiency of QRNG.
\subsubsection{Time Complexity}
Although the original QBS proposal assumes the quantum circuit executes in $O(1)$ time, the depth (longest path) of the circuit does scale with input. The time a quantum circuit takes to execute is the sum of the gate execution times of the gates along the longest path of the circuit, and therefore time complexity scales with circuit depth. \\
For the integer $i$ passed into the QRNG algorithm, we allocate $\lceil \log_2 i \rceil + 1$ qubits. However, this algorithm is rather unique in that gate count and depth don't scale exactly with $i$, but rather \mintinline{python}{bin(i).count('1')}. That is, circuit depth scales with the number of ones in the binary representation of $i$. \\
Consequently, there are large drops in complexity around powers of 2:
\begin{align*}
&11111_2 \text{  (5 ones, high complexity)} \\ &+ 1 = 100000_2 \text{  (1 one, low complexity)}
\end{align*}
For example, $i = 256$ creates a circuit with depth 2 and width 17, but $i = 255$ has a circuit depth of 285 and a width of 17 (perhaps even more depending on transpilation target). This relationship is illustrated in \ref{fig:resources}.
% Also please note that depth + width $\geq$ total.

@@

\subsubsection{Error Rate}
Of course, the main problem with this ballooning circuit size is the fidelity of the measurement; we work with imperfect machines and error rates can compound over time to make a measurement meaningless. 
That being said, this is a QRNG algorithm, so the noise and error generated don't make much of a difference in this scenario (given that the noise is truly random and evenly distributed) besides occasionally measuring a number $ \geq i$. \\
We attempted to create a script to measure the error rate of a circuit based on IBMQ benchmarks, however this only measures the probability of every gate succeeding. This doesn't take into account several errors cancelling each other out or other fortuitous errors, and consequently the measured success rate is astronomically small. A good extension on this would be to use the error rates to modify each gate: For example, an $X$ gate with error $= 1.9765\mathrm{e}{-4}$ would translate to R\textsubscript{x}($(1-1.9765\mathrm{e}{-4}) \cdot \pi$). This would allow for the approximation of the expected position on the Bloch sphere during measurement, thus providing a more accurate error rate.

@@

>>>>

\title{Quantum Bogosort}
\author{George Huebner}% (Grade 12), Michael Caines (Teacher sponsor)}
\date{May 7, 2022}

\begin{document}
\renewcommand{\abstractname}{Preface}
\pagenumbering{gobble}

\begin{titlepage}
\maketitle
\centering
\large Walter Payton College Prep, Chicago IL, United States \\
ISEF Systems Software/Algorithms (SOFT055)% \\
%\large Illinois STEM Fair, Computer Science Research, Design Category \\
%\vspace{3em}
%\huge Sponsor \\
%\vspace{1em}
%\large Michael Caines \hspace{15em}
\end{titlepage}

\pagestyle{fancy}
\fancyhead[R]{Huebner: \textit{Quantum Bogosort} (SOFT055)}
\fancyhead[L]{}

@@

@@

\begin{abstract}
\noindent Joke algorithms are a staple of computer science and provide many a good chuckle. Take, for instance, MiracleSort\textsuperscript{\cite{thompson_2013}}, an algorithm that relies on alpha particle emission to cause erroneous bit flips to sort a list:
\begin{minted}{python}
while isSorted(unsorted_list) is False:
    time.sleep(1000)
\end{minted}
Despite their apparent uselessness, joke algorithms provide valuable insight into algorithm design and complexity\textsuperscript{\cite{gruber_holzer_ruepp_2007}}. In this paper we propose an implementation to Quantum Bogosort, one such joke algorithm.
\end{abstract}

@@

Classic Bogosort is a joke sorting algorithm where a list is randomly permuted and checked to see if it is sorted. In pseudocode:

@@

\noindent It has an average time complexity of $O((n+1)!)$ and the nondeterministic (traditional) method has an unbounded worst case time complexity for any non-trivial list.

Quantum Bogosort\textsuperscript{\cite{the_other_tree_2009}} (QBS), then, is a variation on classical bogosort. Based on the \href{https://www.pbs.org/wgbh/nova/manyworlds/pdf/dissertation.pdf}{many worlds} interpretation of quantum mechanics, QBS proposes that a quantum computer can superimpose all possible permutations of a list and destroy any universe in which a sorted list does not exist, thus sorting the list in $ O(1) $ time. Because parallel universes are non-communicating, there's no way to know the list is sorted other than by the principle of quantum necessity. The algorithm that satisfies subscribers of the Copenhagen interpretation (where wave functions collapse) uses the same algorithm, but the quantum circuit must be run until the list is sorted.

@@

\section{Background}
Most people are likely unfamiliar with quantum computing and its associated quirks, so a minor detour is in order to explain the technologies that will be used to develop and test QBS. This is not meant to be a complete explaination of the quantum mechanics that underlie this project, though, so let the physics enthusiast proceed carefully.

\subsection{Quantum Computers}
Quantum computers should be thought of not as processors that parse instructions, but rather circuits composed of gates that operate on a group of qubits (\textbf{qu}antum \textbf{bit}, the quantum unit of information). While revolutionary, quantum computers only achieve speedups in certain applications, and have not yet cleared a benchmark of true quantum supremacy due to current hardware implementations. Current noisy intermediate scale quantum (NISQ) computers are highly error prone and have comparatively few qubits and are better thought of as a separate processing unit to be orchestrated by a classical computer.
\subsection{Qubits}
Qubits have quantum properties of superposition, entanglement, and interference that distinguish them from classical bits. Whereas a classical bit is either 1 or 0, a qubit is in a superposition of 1 and 0, which are defined as the following kets:
$$ \ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \text{ (Classical 0)} $$
$$ \ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix} \text{ (Classical 1)} $$
Qubit superpositon can be thought of as having a certain amount (amplitude) of `one-ness' and `zero-ness' that can be represented as so:
$$ \ket{x} = \alpha \ket{0} + \beta \ket{1} \ | \ \text{ Probability}_{\ket{0}} = | \alpha |^2 \text{, } \text{ Probability}_{\ket{1}} = | \beta |^2 $$
%Note that $|x|$ here represents magnitude as $\alpha$ and $\beta$ are imaginary.
To actually make use of qubits, they must be measured, which collapses them to either $\ket{0}$ or $\ket{1}$. This is a probabilistic, quantum event, hence the probability\textsubscript{$\ket{x}$}. It is helpful to consider the \hyperref[fig:bsphere]{Bloch sphere} when analyzing qubit state and rotation around the Pauli axes; essentially, the `North Pole' is defined as $\ket{0}$ and the `South Pole' is defined as $\ket{1}$. The `equator,' then, would be an even superposition of $\ket{0}$ and $\ket{1}$.

@@

\noindent Here are a few common qubit gates:
$$ \text{X}(\ket{0}) = \ket{1} $$
$$ \text{X}(\ket{1}) = \ket{0} $$
$$ \text{Z}(\ket{0}) = \ket{0} $$
$$ \text{Z}(\ket{1}) = -\ket{1} $$
$$ \text{H} = \text{R}_{\text{y}} (\frac{\pi}{2}) \text{Z} \ \ \ \text{(Hadamard gate)}$$
$$ \text{H}(\ket{0}) = \frac{1}{\sqrt{2}} \ket{0} + \frac{1}{\sqrt{2}} \ket{1} = \ket{+} $$
$$ \text{H}(\ket{1}) = \frac{1}{\sqrt{2}} \ket{0} - \frac{1}{\sqrt{2}} \ket{1} = \ket{-} $$
Notice that R\textsubscript{y} is an arbitrary rotation around the Pauli Y-axis; Similarly, R\textsubscript{x} and R\textsubscript{z} allow arbitrary rotations along the X- and Z-axes, respectively.\\
Multiple qubits can be used together for entanglement and amplitude interference, and a group of qubits is called a \textit{register}. Although a register can be simplified into a single tensor by taking the tensor product of the qubits in the register, keeping registers in kets makes the math much simpler. Bra-ket (Dirac) notation is used as Kronecker products quickly become unwieldy when dealing with large qubit registers.

@@

\subsection{Entanglement}
Qubit gates can also be controlled if another qubit is $\ket{1}$ (similar to an `if' clause in a classical programming language). This allows for multiple qubits to be entangled with each other, essentially tying the 'fate' of the qubits together so that they can no longer be thought of as individual qubits but rather a single system. Measuring one qubit will also collapse the other entangled qubits, even if they were never observed. A famous example of this is Schr\"odinger's cat. In this thought experiment, the quantum event of a radioactive substance decaying is entangled with the fate of the cat. If the cat dies, the radioactive substance must have decayed, and if the radioactive substance did not decay, the cat must be alive (or, at the very least, not killed by poison).

@@

\item The other approach explored is essentially a quantum random number generator. Instead of storing the list, we create $N = l!$ evenly balanced register states ($l!$ being the number of permutations of $array$). A set of states $[0, N) \cap \mathbb{N}$ will be prepared, each with an even amplitude. Concisely:
    $$ \sum_{x=0}^{N-1} \frac{1}{\sqrt{N}} \ket{x_{BE}} $$
Note that the quantum register will always be represented in \href{https://en.wikipedia.org/wiki/Endianness}{big-endian}.

Because we're storing our number in binary, we want the power of 2 greater than or equal to $l!$. This means we have $O(\lceil \log_2 l! \rceil)$ space complexity, which seems scary at first. However, with present-day NISQ technology, this is significantly more favorable than $O(l \cdot s)$ for any useful data type. Also keep in mind that because element size isn't a concern, any size element or even elements of different types could be stored in the list (as long as every type used can be compared to each other).

@@

\section{Quantum Bogosort}
\subsection{Quantum Random Number Generation (QRNG) Algorithm}
The main algorithm is recursive, essentially moving through the register from most to least significant qubit, performing a R\textsubscript{y} gate to split up a qubit into P 'parts,' and entangling the remaining qubits with controlled Hadamard (CH) gates. \\
Let's use $i = 77$ as an example. Recall that for storing an integer $i$, the most significant bit (MSB) used is $\lfloor \log_2 i \rfloor$: For example, the MSB of 77 would be $\lfloor \log_2 77 \rfloor = 64$. Because QRNG generates the integers $[0, 77)$ (excluding 77 itself in order to generate 77 integers), we know that $[0, 64)$ is also being generated, because $[0, 64) \subseteq [0, 77)$. \\
This means that every possible state of the qubit register with a leading zero (in this case, in the 64s place) should be generated. That's already really easy to do! Just use the leading zero as a control and apply Hadamard gates to the rest of the register. Here's an example:
\begin{align*}
& \ket{\psi} = \ket{0000000} \\
& H(\psi_0), \ \ \ \ \ \ \ket{\psi} = \frac{1}{\sqrt{2}} \ket{0000000} + \frac{1}{\sqrt{2}} \ket{1000000} \\
& CH(\psi_0, \psi_{[1, 6]}), \ X(\psi_0), \ \ \ \ \ \ \ \ket{\psi} = \frac{1}{\sqrt{128}}(\ket{0000000} + \ket{0000001} + ... + \ket{0111111}) + \frac{1}{\sqrt{2}} \ket{1000000}
\end{align*}
Note that CH is a singly controlled Hadamard applied to multiple qubits, not a multi-controlled Hadamard applied to one qubit. Also note the X gate applied to the MSB because we really want an anti-controlled, not a normal controlled gate. \\
The problem is that we've unevenly split the integer 77 into 2: $|[0, 64)| \neq |[64, 77)|$. Numbers in the range $[64, 77)$ have a $(\frac{1}{\sqrt{13}} \cdot \frac{1}{\sqrt{2}})^2$ probability of being measured whereas numbers in the $[0, 64)$ range have a $(\frac{1}{\sqrt{64}} \cdot \frac{1}{\sqrt{2}})^2$ probability of being measured. \\
To fix this, we have to change the $\frac{1}{\sqrt{2}}$ from the Hadamard gate to something that gives us a more complex superposition; to do this we use the arbitrary rotation gates. $X$ and $Y$ gates give us the desired rotations we want, but we chose $Y$ so we don't have to worry about phase (the sign of a qubit). Phase doesn't matter in this case because only the magnitudes effect the probability of measurement, but it might be important if using QRNG in another algorithm. \\
Superpositions can be represented using spherical coordinates. For example, on the \hyperref[fig:bsphere2]{Bloch sphere}, the spherical coordinates $(1, 0.955, 0)$ would represent the state $\sqrt{\frac{1}{3}} \ket{0} + \sqrt{\frac{2}{3}} \ket{1}$ because $\cos (0.955) \approx \sqrt{\frac{1}{3}}$. We can do some algebra to get a $\theta$ for an arbitrary superposition. For QRNG, this looks like $\theta = 2\arccos(\sqrt{\frac{77 - MSB}{77}})$. \\
\begin{figure}[H]
    \centering
    \capstart
    \includegraphics[width=0.19\textwidth]{images/bloch_sphere2.png}
    \caption{Bloch sphere\textsuperscript{\cite{wikipedia_2021}}}
    \label{fig:bsphere2}
\end{figure}
\noindent Recall that we have to apply a $X$ gate because gates are controlled by $\ket{1}$, not $\ket{0}$. We can save an extra gate by doubling our rotation (hence the 2), which gives us $\sqrt{\frac{77 - MSB}{77}} \ket{0}$ instead, which flips right back to $\ket{1}$ with another $X$ gate. \\
\indent We then recursively repeat this process, adding the MSB as a control qubit to the R\textsubscript{y} and CH gates and using the 2nd most significant bit as the new MSB. Eventually we split all the states evenly and we're done. \\

\noindent The differences between what is conceptually happening and what is actually implemented are shown below. $\bullet$ is a control and $\circ$ is an anti-control (controls on $\ket{0}$ instead of $\ket{1}$). Note that H\textsubscript{ALL} is \textbf{not} a multi-qubit gate --- it's illustrated that way for clarity.
\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{images/circuits/conceptual_circuit.pdf}
    \caption{Conceptual circuit\textsuperscript{\cite{chuang}}}
    \vspace{10mm}
    \includegraphics[width=\textwidth]{images/circuits/optimized_circuit.pdf}
    \caption{Circuit with proper controls\textsuperscript{\cite{chuang}}}
    \label{fig:circs}
\end{figure}

\vspace{10mm}
\noindent Minor implementation details:
\begin{itemize}
    \item Multi-controlled Hadamard (MCH) gates are implemented manually with an ancilla qubit and multi-controlled X (MCX) gates. This could almost certainly be optimized more on a case-by-case basis (i.e. running \mintinline{python}{mode='noancilla'} for machines with fewer qubits).
    \item Especially when there are several qubits in the control register (and therefore the MCH gates are more costly), the MCX on the ancilla doesn't have to be reversed; instead, applying a  \mintinline{python}{'reset'} gate to reset the ancilla to 0 might be more optimal.
    \item \mintinline{python}{circuit} and \mintinline{python}{ancilla} don't change between recursions and needlessly add on to the stack frame of the function. Passing them as parameters is intended for clarity, but if you were to squeeze every drop of performance out they should be made global.
\end{itemize}
\subsubsection{Code}
\begin{minted}{python}
from qiskit import *
from math import acos, sqrt

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{images/example_circuit.png}
    \caption{Circuit diagram for 15 `parts'}
    \label{fig:circ_diagram}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.75\textwidth]{images/example_measurement.png}
    \caption{Measurement results from Aer Simulator}
    \label{fig:measures}
\end{figure}

\subsection{Practicality Assessment}
Obviously bogosort is not meant to be a practical algorithm, but nevertheless we will conduct a review of the efficiency of QRNG.
\subsubsection{Complexity}
Although the original QBS proposal assumes the quantum circuit executes in $O(1)$ time, the depth (longest path) of the circuit does scale with input. The time a quantum circuit takes to execute is the sum of the gate execution times of the gates along the longest path of the circuit, and therefore time complexity scales with circuit depth. \\
For the integer $i$ passed into the QRNG algorithm, we allocate $\lceil \log_2 i \rceil + 1$ qubits. However, this algorithm is rather unique in that gate count and depth don't scale exactly with $i$, but rather \mintinline{python}{bin(i).count('1')}. That is, circuit depth scales with the number of ones in the binary representation of $i$. \\
Consequently, there are large drops in complexity around powers of 2:
\begin{align*}
&11111_2 \text{  (5 ones, high complexity)} \\ &+ 1 = 100000_2 \text{  (1 one, low complexity)}
\end{align*}
For example, $i = 256$ creates a circuit with depth 2 and width 17, but $i = 255$ has a circuit depth of 285 and a width of 17 (perhaps even more depending on transpilation target). This relationship is illustrated in \ref{fig:resources}.
% Also please note that depth + width $\geq$ total.

@@

\subsubsection{Error Rate}
Of course, the main problem with this ballooning circuit size is the fidelity of the measurement; we work with imperfect machines and error rates can compound over time to make a measurement meaningless. 
That being said, this is a QRNG algorithm, so the noise and error generated don't make much of a difference in this scenario (given that the noise is truly random and evenly distributed) besides occasionally measuring a number $ \geq i$. \\
%We attempted to create a script to measure the error rate of a circuit based on IBMQ benchmarks, however this only measures the probability of every gate succeeding. This doesn't take into account several errors cancelling each other out or other fortuitous errors, and consequently the measured success rate is astronomically small. A good extension on this would be to use the error rates to modify each gate: For example, an $X$ gate with error $= 1.9765\mathrm{e}{-4}$ would translate to R\textsubscript{x}($(1-1.9765\mathrm{e}{-4}) \cdot \pi$). This would allow for the approximation of the expected position on the Bloch sphere during measurement, thus providing a more accurate error rate.
One way we can model error is by wrapping our gates with error gates --- identity gates with an error rate corresponding to a noise profile sampled from real quantum hardware. This approach works with gate errors, including measurement and reset errors, and could also be applied to decoherence and relaxation errors. \\
Often times noise on hardware will turn a pure quantum state into a mixed state, a composition of pure states with classical probabilities (which is \textbf{not} the same as superposition); for example, perhaps an R\textsubscript{Y}$(\frac{\pi}{2})$ will succeed $80\%$ of the time and over/undershoot by $\frac{\pi}{6}$ 20\% of the time. We can write this mixed state as
\begin{align*}
\ket{\psi_0} &= \frac{1}{\sqrt{2}} \left( \ket{0} + \ket{1} \right) \ \ \ \text{(80\% chance)} \\
\ket{\psi_1} &= \frac{1}{\sqrt{4}} \ket{0} + \sqrt{\frac{3}{4}} \ket{1} \ \ \ \text{(10\% chance)} \\
\ket{\psi_2} &= \sqrt{\frac{3}{4}} \ket{0} + \frac{1}{\sqrt{4}} \ket{1} \ \ \ \text{(10\% chance)}
\end{align*}
Density matrices are particularly convenient because we can represent this mixed state in a single matrix:
\begin{align*}
    \rho &= \frac{4}{5} \ket{\psi_0}\bra{\psi_0} + \frac{1}{10} \ket{\psi_1}\bra{\psi_1} + \frac{1}{10} \ket{\psi_2}\bra{\psi_2} \\
    &= \frac{4}{5} \begin{bmatrix} \frac{1}{2} & \frac{1}{2} \\ \frac{1}{2} & \frac{1}{2} \end{bmatrix} + \frac{1}{10} \begin{bmatrix} \frac{1}{4} & \frac{\sqrt{3}}{4} \\ \frac{\sqrt{3}}{4} & \frac{3}{4} \\ \end{bmatrix} + \frac{1}{10} \begin{bmatrix} \frac{3}{4} & \frac{\sqrt{3}}{4} \\ \frac{\sqrt{3}}{4} & \frac{1}{4} \\ \end{bmatrix} \\
    &= \begin{bmatrix} \frac{1}{2} & \frac{2}{5} + \frac{\sqrt{3}}{20} \\ \frac{2}{5} + \frac{\sqrt{3}}{20} & \frac{1}{2} \\ \end{bmatrix} \ \ \text{Purity: 0.9736}
\end{align*}
In the case of quantum bogosort, we can calculate the density matrix for the idealized ($\rho$) and noisy ($\rho'$) circuits; Both are transpiled for IBMQ Santiago and the latter is injected with Santiago's noise profile. $\rho'$'s purity of 0.42118 reflects that noise heavily affects the fidelity of the qubits.
\begin{gather*}
    \rho = \begin{bmatrix}
        \frac{1}{15} & \frac{1}{15} & \cdots & \frac{1}{15} & 0 \\
        \frac{1}{15} & \frac{1}{15} & \cdots & \frac{1}{15} & 0 \\
        \vdots & \vdots & \ddots & \vdots & \vdots \\
        \frac{1}{15} & \frac{1}{15} & \cdots & \frac{1}{15} & 0 \\
        0 & 0 & \cdots & 0 & 0 \\
    \end{bmatrix} \\
    \rho' = \begin{bmatrix}
        0.06997 & 0.04275 - 0.00057i & \cdots & 0.0376 + 0.00028i & 0.00036 + 0.00024i  \\
        0.04275 + 0.00057i & 0.06646 & \cdots & 0.03687 + 0.00025i & 0.00002 + 0.0001i  \\
        \vdots & \vdots & \ddots & \vdots & \vdots \\
        0.0376 - 0.00028i & 0.03687 - 0.00025i & \cdots & 0.06272 & 0.00217 + 0.00045i  \\
        0.00036 - 0.00024i & 0.00002 - 0.0001i & \cdots & 0.00217 - 0.00045i & 0.01268  \\
    \end{bmatrix}
\end{gather*}

@@

\pagebreak
